---
title: "Air Quality North Yorkshire"
author: "Olasho"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    highlight: zenburn
    theme: cerulean
    number_sections: true
---

# Document specification

```{r}
knitr::opts_chunk$set(comment = " ", warning = FALSE, message = "FALSE", fig.show = "hold")

options(knitr.table.format = "html")
```


```{r, results='hide', warning=FALSE, message=FALSE}
 #install.packages("xlsx")
library(tidyverse)
library(sf)
library(kableExtra)
library(stringr)
library(htmlTable)
library(tableone)
library(tmap)
library(paletteer)
library(ggthemes)
library(readxl)

# loading the lsoa shapefile

lsoa_deliverble <- st_read("X:/HDRC/Workstream - Data/Air quality/CERC_WYCA_work/Data/LSOA_Deliverable_i2.shp")

WestYorkshire_500mgrid <- st_read("X:/HDRC/Workstream - Data/Air quality/CERC_WYCA_work/Data/WestYorkshire_500mgrid.SHP")

# inspecting the lsoa_deliverable

head(lsoa_deliverble)

glimpse(lsoa_deliverble)

list(unique(lsoa_deliverble$LA))

# inspecting the WestYorkshire_500mgrid

head(WestYorkshire_500mgrid)
```

# Selecting the variables used

```{r}

columns_to_remove <- c(9:16, 18:24, 28, 35:37, 39, 42:49, 51:57, 61, 68:70, 72, 75:82, 84:90, 94, 101:103, 105)

lsoa_deliverble_1 <- lsoa_deliverble[ , -columns_to_remove]

st_crs(lsoa_deliverble_1)
```

# Descriptive analysis

Here, we are going to carry out summary statistics for the pollutants.v

## PM25

```{r, fig.show='hold'}
# First drop the geometry

lsoa_deliverble_2 <- lsoa_deliverble_1 %>%
  st_drop_geometry()

# Extract the PM25 variables

PM25_dataset <- lsoa_deliverble_2 %>%
  select(LA, starts_with("PM25"))

# Creating an empty dataframe

PM25_summary_table <- data.frame(pollutant = character(), mean = numeric(), sd = numeric(), minimum = numeric(), maximum = numeric(), stringsAsFactors = FALSE)

# Define the pollutants

pollutants <- c("PM25_02a", "PM25_02b", "PM25_01", "PM25_03", "PM25_04", "PM25_08a", "PM25_08b", "PM25_08c", "PM25_08d", "PM25_08e", "PM25_08", "PM25_HB", "PM25", "PM25_km2")

# Using a for loop to fill the empty dataframe with PM25

for (pollutant in pollutants) {
  # Calculate statistics for each pollutant
  pollutant_stats = PM25_dataset %>%    
    summarise(mean = mean(get(paste0(pollutant))),
              sd = sd(get(paste0(pollutant))),
              minimum = min(get(paste0(pollutant))),
              maximum = max(get(paste0(pollutant)))
              ) %>% 
    mutate(PM25_PM10_NOx_source = pollutant)
  # Bind the results to the empty dataframe
  PM25_summary_table <- rbind(PM25_summary_table, pollutant_stats)
}

# Moving the pollutant column to the begining of the data frame

PM25_summary_table <- PM25_summary_table %>%
  select(PM25_PM10_NOx_source, everything())

# Renaming the pollutants 

## Define a vector of the new column

PM25_column_name <- c("Total Dometic combustion", "total commercial combustion", "Energy production", "Industrial combustion", "Industrial processes", "Aircraft transport", "Rail transport", "Industrial off-road mobile machinery", "Other off-road mobile machinery", "Transport residual", "Transport total", "Heather burning", "Total all sectors", "Weighted PM25")

## Rename the pollutant columns

PM25_summary_table$PM25_PM10_NOx_source <- PM25_column_name

## Roundingup the table to 3 decimal places

PM25_summary_table[ , 2:5] <- round(PM25_summary_table[ , 2:5], digits = 3)

PM25_summary_table
```

## PM10

```{r, fig.show='hold'}
# Extract the PM25 variables

PM10_dataset <- lsoa_deliverble_2 %>%
  select(LA, starts_with("PM10"))

# Creating an empty dataframe

PM10_summary_table <- data.frame(pollutant = character(), mean = numeric(), sd = numeric(), minimum = numeric(), maximum = numeric(), stringsAsFactors = FALSE)

# Define the pollutants

pollutants <- c("PM10_02a", "PM10_02b", "PM10_01", "PM10_03", "PM10_04", "PM10_08a", "PM10_08b", "PM10_08c", "PM10_08d", "PM10_08e", "PM10_08", "PM10_HB", "PM10")

# Using a for loop to fill the empty dataframe with PM25

for (pollutant in pollutants) {
  pollutant_stats = PM10_dataset %>%
    summarise(mean = mean(get(paste0(pollutant))),
              sd = sd(get(paste0(pollutant))),
              minimum = min(get(paste0(pollutant))),
              maximum = max(get(paste0(pollutant)))
              ) %>% 
    mutate(PM25_PM10_NOx_source = pollutant)
  PM10_summary_table <- rbind(PM10_summary_table, pollutant_stats)
}

# Moving the pollutant column to the beginning of the data frame

PM10_summary_table <- PM10_summary_table %>%
  select(PM25_PM10_NOx_source, everything())

# Renaming the pollutants 

## Define a vector of the new column

PM10_column_name <- c("Total Dometic combustion", "total commercial combustion", "Energy production", "Industrial combustion", "Industrial processes", "Aircraft transport", "Rail transport", "Industrial off-road mobile machinery", "Other off-road mobile machinery", "Transport residual", "Transport total", "Heather burning", "Total all sectors")

## Rename the pollutant columns

PM10_summary_table$PM25_PM10_NOx_source <- PM10_column_name

## Roundingup the table to 3 decimal places

PM10_summary_table[ , 2:5] <- round(PM10_summary_table[ , 2:5], digits = 3)

PM10_summary_table
```

## NOx 

```{r, fig.show='hold'}
# Extract the PM25 variables

NOx_dataset <- lsoa_deliverble_2 %>%
  select(LA, starts_with("NOx"))

# Creating an empty dataframe

NOx_summary_table <- data.frame(pollutant = character(), mean = numeric(), sd = numeric(), minimum = numeric(), maximum = numeric(), stringsAsFactors = FALSE)

# Define the pollutants

pollutants <- c("NOx_02a", "NOx_02b", "NOx_01", "NOx_03", "NOx_04", "NOx_08a", "NOx_08b", "NOx_08c", "NOx_08d", "NOx_08e", "NOx_08", "NOx_HB", "NOx")

# Using a for loop to fill the empty dataframe with PM25

for (pollutant in pollutants) {
  pollutant_stats = NOx_dataset %>%
    summarise(mean = mean(get(paste0(pollutant))),
              sd = sd(get(paste0(pollutant))),
              minimum = min(get(paste0(pollutant))),
              maximum = max(get(paste0(pollutant)))
              ) %>% 
    mutate(PM25_PM10_NOx_source = pollutant)
  NOx_summary_table <- rbind(NOx_summary_table, pollutant_stats)
}

# Moving the pollutant column to the beginning of the data frame

NOx_summary_table <- NOx_summary_table %>%
  select(PM25_PM10_NOx_source, everything())

# Renaming the pollutants 

## Define a vector of the new column

NOx_column_name <- c("Total Dometic combustion", "total commercial combustion", "Energy production", "Industrial combustion", "Industrial processes", "Aircraft transport", "Rail transport", "Industrial off-road mobile machinery", "Other off-road mobile machinery", "Transport residual", "Transport total", "Heather burning", "Total all sectors")

## Rename the pollutant columns

NOx_summary_table$PM25_PM10_NOx_source <- NOx_column_name

## Roundingup the table to 3 decimal places

NOx_summary_table[ , 2:5] <- round(NOx_summary_table[ , 2:5], digits = 3)

NOx_summary_table

# Binding all three tables of PM25, PM10, and NOx together

PM25_PM10_NOX_summary_table <- bind_rows(PM25_summary_table, PM10_summary_table, NOx_summary_table)
```

### Presentation of tables in standard format

```{r, fig.show='hold'}
PM25_PM10_NOX_summary_table %>%
  kbl(caption = "Summary table of air pollutants by emmision source") %>%
  kable_paper("hover") %>%
  kable_classic_2(full_width = F, html_font = "Cambria")
```

# Summary statistics by local Authority

```{r, fig.show='hold'}
View(lsoa_deliverble_1)
table(lsoa_deliverble_1$LA, lsoa_deliverble_1$PM25_02b)
```


# Spatial analysis
## By LSOA

First, spatial analysis of the geographical distribution of the total Tonnes of PM25, NOx, PM10 by LSOA was carried out.

```{r, fig.show='hold'}
# First subset the lsoa_deliverable_1 to just include LSOA and the total PM25

PM25_LSOA_PMTotal <- lsoa_deliverble_1 %>%
  select(LSOA21CD, LSOA21NM, LA, PM25)

# Using tmap to plot the map of total PM25 distribution by LSOA

PM25_LSOA_PMTotal_map <- tm_shape(PM25_LSOA_PMTotal) +
  tm_fill("PM25", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 2) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Annual total tonnes of PM25 by lSOA") +
  tm_borders(alpha = 0.5,
             lwd = 2)
  
# subset the lsoa_deliverble_1 to just include LSOA and the total PM10

PM10_LSOA_PMTotal <- lsoa_deliverble_1 %>%
  select(LSOA21CD, LSOA21NM, LA, PM10)

# Using tmap to plot the map of total PM25 distribution by LSOA

PM10_LSOA_PMTotal_map <- tm_shape(PM10_LSOA_PMTotal) +
  tm_fill("PM10", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 10) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Annual total tonnes of PM10 by LSOA") +
  tm_borders(alpha = 0.5,
             lwd = 2)

# subset the lsoa_deliverble_1 to just include LSOA and the total NOx

NOx_LSOA_PMTotal <- lsoa_deliverble_1 %>%
  select(LSOA21CD, LSOA21NM, LA, NOx)

# Using tmap to plot the map of total NOx distribution by LSOA

NOx_LSOA_PMTotal_map <- tm_shape(NOx_LSOA_PMTotal) +
  tm_fill("NOx", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 10) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Annual total tonnes of NOx by LSOA") +
  tm_borders(alpha = 0.5,
             lwd = 2)

# Viewing all three maps

tmap_mode("view")
a <- tmap_arrange(PM25_LSOA_PMTotal_map, PM10_LSOA_PMTotal_map, NOx_LSOA_PMTotal_map, nrow = 2)
a
```

## Spatial facet by LA of PM25, NOx, and PM10

Here, we are going to split the map by LA to have a clearer view of the pollutant's distribution by LSOA for each local AUthority.

```{r, fig.show='hold'}

# PM25

PM25_LSOA_facet <- lsoa_deliverble_1
  
# Plotting the map

tmap_mode("plot")
PM25_LSOA_facet_map <- tm_shape(PM25_LSOA_facet) +
  tm_fill("PM25", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 2,
          legend.hist = TRUE,
          thres.poly = 0) +
  tm_layout(title.size = 20,
            title = "PM25") +
  tm_borders(alpha = 0.5,
             lwd = 2) +
  tm_facets(by = "LA", free.scales = FALSE,
            free.coords = TRUE,
            drop.units = TRUE) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(widt =0.2)

PM25_LSOA_facet_map
```

```{r, fig.show='hold'}
# PM10

## Plotting the map

tmap_mode("plot")
PM10_LSOA_facet_map <- tm_shape(lsoa_deliverble_1) +
  tm_fill("PM10", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 2,
          legend.hist = TRUE,
          thres.poly = 0) +
  tm_layout(title.size = 20,
            title = "PM10") +
  tm_borders(alpha = 0.5,
             lwd = 2) +
  tm_facets(by = "LA", free.scales = FALSE,
            free.coords = TRUE,
            drop.units = TRUE) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(widt =0.2)

PM10_LSOA_facet_map
```

```{r, fig.show='hold'}
# NOx

## Plotting the map

tmap_mode("plot")
NOx_LSOA_facet_map <- tm_shape(lsoa_deliverble_1) +
  tm_fill("NOx", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 2,
          legend.hist = TRUE,
          thres.poly = 0) +
  tm_layout(title.size = 20,
            title = "NOx") +
  tm_borders(alpha = 0.5,
             lwd = 2) +
  tm_facets(by = "LA", free.scales = FALSE,
            free.coords = TRUE,
            drop.units = TRUE) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(widt =0.2)

NOx_LSOA_facet_map
```

## By Local Authority

Spatial analysis of the geographical distribution of the average PM25, NOx, PM10 by Local Authority was carried out.

```{r, fig.show='hold'}
# PM25

PM25_LA <- lsoa_deliverble_1 %>%
  group_by(LA) %>%
  summarise(mean_PM25 = mean(PM25))

PM25_LA_map <- tm_shape(PM25_LA) +
  tm_fill("mean_PM25", 
          style = "kmeans", 
          palette = paletteer_c("ggthemes::Orange", 30), 
          legend.is.portrait = TRUE,
          fill_width = 10) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Average PM25 by LA") +
  tm_borders(alpha = 0.5,
             lwd = 3) +
  tm_text("LA")

# PM10

PM10_LA <- lsoa_deliverble_1 %>%
  group_by(LA) %>%
  summarise(mean_PM10 = mean(PM10))

PM10_LA_map <- tm_shape(PM10_LA) +
  tm_fill("mean_PM10", 
          style = "kmeans", 
          palette = paletteer_c("ggthemes::Orange", 30), 
          legend.is.portrait = TRUE,
          fill_width = 10) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Average PM10 by LA") +
  tm_borders(alpha = 0.5,
             lwd = 3) +
  tm_text("LA")

# NOx

NOx_LA <- lsoa_deliverble_1 %>%
  group_by(LA) %>%
  summarise(mean_NOx = mean(NOx))

NOx_LA_map <- tm_shape(NOx_LA) +
  tm_fill("mean_NOx", 
          style = "kmeans", 
          palette = paletteer_c("ggthemes::Orange", 30), 
          legend.is.portrait = TRUE,
          fill_width = 10) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Average NOx by LA") +
  tm_borders(alpha = 0.5,
             lwd = 3) +
  tm_text("LA")

tmap_mode("view")
tmap_arrange(PM25_LA_map, PM10_LA_map, NOx_LA_map, nrow = 2)
```

######## Spatial investigation of the source of emmision contributing the most to poor air quality by local authority

Here, we are going to investigate the source of emmision that contributes to most to the pollutants, by overlaying the clusters with the average pollutants produced by the source of emission which includes, total pollutants by domestic combustion, commercial combustion, transport, and heather burning.

```{r, fig.show='hold'}
# PM25
## preparing the data

PM25_overlay_data <- lsoa_deliverble_1 %>%
  group_by(LA) %>%
  summarise(mean_PM25 = mean(PM25), mean_domestic = mean(PM25_02a), mean_commercial = mean(PM25_02b), mean_transport = mean(PM25_08), mean_heather = mean(PM25_HB))

PM_overlay_result <- st_intersection(PM25_overlay_data)

## Plotting with tmap
tm_shape(PM25_overlay_data) +
  tm_dots(size = "mean_PM25", col = "red") +
  tm_shape(PM25_overlay_data) +
  tm_dots(size = "mean_domestic", col = "green") +
  tm_dots(size = "mean_commercial", col = "blue") +
  tm_dots(size = "mean_transport", col = "yellow") +
  tm_dots(size = "mean_heather", col = "black") +
  tm_borders(alpha = 0.5, lwd = 2)
```

## Spatial analysis of the Bradford district


```{r}

shp_file <- st_read("C:/Users/Shobande1/OneDrive - bradford.gov.uk/Documents/Bradford/R language/UK_shapefile/WD_DEC_2022_UK_BFC.shp")

```

## Comparing the correlation between Poor air quality and Standardized Mortality Ratio (SMR) of Respiratory disease

Here we are going to compare the spatial correlation of poor air quality with respiratory disease SMR.

```{r, fig.show='hold'}
# Importation of the SMR dataset

Resp_SMR <- read_xlsx("C:/Users/Shobande1/OneDrive - bradford.gov.uk/Documents/Bradford/Excel sheets/Resp_SMR.xlsx")

# Joining the Resp_SMR data with the shapefile

Resp_shp <- left_join(shp_file, Resp_SMR, by = c("WD22CD" = "Code"))

# Selecting all the LA in West Yorkshire

Resp_shp_WY <- Resp_shp %>%
  filter(LAD22NM == "Bradford" | LAD22NM == "Leeds" | LAD22NM == "Kirklees" | LAD22NM == "Calderdale" | LAD22NM == "Wakefield")
View(Resp_shp_WY)
# Checking for validity of the spatial object

st_is_valid(Resp_shp_WY)

# From the result above, there appears to be one invalid geometry. To fix, use st_make_valid

Resp_shp_WY_2 <- st_make_valid(Resp_shp_WY)

######################################################## OR

Resp_shp_WY_3 <- st_simplify(Resp_shp_WY)

# plotting the spatial analysis of Respiratory SMR

Resp_shp_WY_map <- tm_shape(Resp_shp_WY_2) +
  tm_fill("Deaths from respiratory diseases, all ages", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 2,
          title = "Respiratory disease SMA by Ward") +
  tm_borders(alpha = 0.5,
             lwd = 2)

tmap_mode("view")
Resp_shp_WY_map

# Spltting the plot by local authority

Resp_shp_WY_map_facet <- tm_shape(Resp_shp_WY) +
  tm_fill("Deaths from respiratory diseases, all ages", 
          style = "kmeans", 
          palette = paletteer_c("grDevices::Plasma", 30), 
          legend.is.portrait = TRUE,
          fill_width = 2,
          legend.hist = TRUE,
          thres.poly = 0) +
  tm_layout(title.size = 20,
            title = "Respiratory disease SMR") +
  tm_borders(alpha = 0.5,
             lwd = 2) +
  tm_facets(by = "LAD22NM", free.scales = FALSE,
            free.coords = TRUE,
            drop.units = TRUE) +
  tm_compass(type = "8star", size = 2) +
  tm_scale_bar(width =0.2)

tmap_mode("plot")
Resp_shp_WY_map_facet

# Finding the average Resp SMA mean by LA

Resp_SMA_LA <- Resp_shp_WY %>%
  group_by(LAD22NM) %>%
  summarise(mean_SMA = mean(`Deaths from respiratory diseases, all ages`))

# plotting the tmap

Resp_SMA_LA_map <- tm_shape(Resp_SMA_LA) +
  tm_fill("mean_SMA", 
          style = "kmeans", 
          palette = paletteer_c("ggthemes::Orange", 30), 
          legend.is.portrait = TRUE,
          fill_width = 10) +
  tm_layout(legend.height = 1,
            legend.outside.size = 1,
            legend.outside = TRUE,
            legend.position = c("right", "top"),
            frame = TRUE,
            title = "Average Respiratory Disease SMA by LA") +
  tm_borders(alpha = 0.5,
             lwd = 3) +
  tm_text("LAD22NM")

tmap_mode("view")
Resp_SMA_LA_map

# viewing the map alongside the air pollutants

b <- tmap_arrange(PM25_LA_map, PM10_LA_map, NOx_LA_map, Resp_SMA_LA_map, nrow = 2)
```


```{r}
tmap_save(Resp_SMA_LA_map, filename = "C:/Users/Shobande1/Desktop/Resp_SMA_LA_map.html", dpi = 3, height = 1000, width = 1000, units = "px", filetype = "html")
```

